Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-01-24 manu_draft.Rmd
methods revision from JH done

Next we consider the goal of estimating a seasonal average, such as the mean spring chlorophyll concentrations and the estimated uncertainty in each year.  Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  For simplicity, the following text omits subscript $t$.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.  Here, the seasonal averages used in our examples are calculated at the resolution of days.  The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.  The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.  Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.  This is given by $\hShy = \bX \hShb \bX^T$.

Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.  Each of these estimates are standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.  A similar summary for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series is provided by @Murphy19.  For example, the above estimates can be used for a test of the null hypothesis that `y` is the same at two times. To do so, $\bX$ would have two rows for each time to compare.  The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  This difference can then be used to determine if the two estimates in time are significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.

@@

This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.  In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.  These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the SFE region. For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:

1. Do two years, or a group of years, differ significantly in a seasonal average?
2. Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?  This question can be posed in a moving-window manner across a time-series.

For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.

### Differences among years in seasonal averages

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].  While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.

### Linear trends across years at a chosen time-scale of interest

The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19]:

@@

where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$. Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.  The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.  The residuals is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.  The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.  The random effect is assumed a normal distribution with mean 0 and unknown variance, $\sbs$, to be estimated.

The R package *mixmeta* was used to estimate the model in equation \@ref(eq:mixmet) [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).  The default estimation method for meta-analysis models is restricted maximum likelihood (REML).

We apply this method for windows of a chosen width of years.  This approach evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).  Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error. The meta-analysis explicitly includes these differences in uncertainty magnitudes.

As a final analysis, the chosen windows for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not.  Although the initial window width and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year.  Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

>>>>

In the second-stage analysis, we estimated a seasonal average, such as the mean spring chlorophyll concentrations and the uncertainty in each year.  Define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  For simplicity, the following text omits subscript $t$.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.  Here, the seasonal averages used in our examples were calculated at the resolution of days.  The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.  The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$, wheras the off-diagonal elements are the correlations among the elements of $\hbb$.  Since parameter estimates are correlated, the covariance of $\hby$ is $\hShy = \bX \hShb \bX^T$.

The estimated seasonal average was calculated from the vector of daily values for each of the $n$ days in the season of interest with $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.  Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$. @Murphy19 provided a similar approach for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series.  For example, the above estimates were used for a test of the null hypothesis that `y` is the same at two times. To do so, $\bX$ would have two rows for each time to compare.  The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  This difference was used to determine if the two estimates in time were significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates. An alternative interpretation of this test is a quantification of differences between two points in time as affected by the magnitude of the difference and range of likely values for the magnitude of change.

@@

In stage three of the analysis, we used a meta-analysis method to evaluate across-year trends of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.  In the examples, we considered two periods of interest for chlorophyll, January-July and August-December, which are relevant to phytoplankton bloom phenology in south SFE.

This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?  For example, is there a trend in the spring chlorophyll average from 1990 to 2000? This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.). For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.

### Linear trends across years at a chosen time-scale

A meta-analysis mixed effects model is useful to estimate linear trends when each observation has a unique and quantified standard error, which is the case with our estimates $\hmt$ and $\hshmt$.  Differences in standard errors, which may result from different monitoring effort between years, are explicitly considered in the analysis.  The model can be expressed using notation similar to @Sera19:

@@

where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$. Accordingly, the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.  The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.  The residuals are assumed to be independent and normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.  The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.  The random effect is assumed a normal distribution with mean 0 and unknown variance, $\sbs$, to be estimated.

We estimated the model (equation \@ref(eq:mximet)) using the *mixmeta* package in R [@Sera19]. Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed and others estimated ($\sbs$).  The default estimation method for meta-analysis models is restricted maximum likelihood (REML).

This method was applied to a chosen sequence or "window" of years for estimating the linear trend.  This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chlorophyll over a chosen series of years. Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).  Years with more data have smaller standard errors for $\hmt$ compared to years with less data. The meta-analysis explicitly includes these differences in uncertainty magnitudes.

As a final analysis, the chosen sequence of years for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).  Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year (e.g., an estimate for 1990 for a six year window assesses the trend from 1987 to 1993).  Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
